---
title: Modifying the input structure 
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Modifying the input structure}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

# avoid check where vignette filename must equal the vignette title
options(rmarkdown.html_vignette.check_title = FALSE)
```


```{r setup, include = FALSE}
pkg = 'flowdiagramr' #so we can easily switch names
library(pkg, character.only = TRUE)
```

# Introduction

We assume you went through the [getting started vignette](./A_quickstart.html) and also the [basic modification vignette](./B_modify_diagrams_simple.html). 

If the basic modifications you can achieve by providing different options to `make_diagram`, then you do not need to go any further, and there is no need to understand in more detail how **`r pkg`** works. However, it is quite likely that at some point, you want to further tweak your diagram. To do so successfully, it is useful to learn some more details about how the package works. 

As you learned, producing a diagram with **`r pkg`** involves two steps. After you have defined your model, there is a first step that takes your model and creates a list of data frames (which we called the `input_structure` in the prior vignettes). These data frames are used by `make_diagram` to produce the ggplot code and figure. 

The advantage of this 2-step process is that you can make modifications to the `input_structure` object before sending it into the `make_diagram` function, and thus modify the layout of the diagram beyond what you can through the`make_diagram` options, while still being fairly easy to do. 



# Explanation of the `make_diagram` input object

To modify the object returned from `prepare_diagram` and used as input for `make_diagram`, it is important to learn a bit about it.
You can learn the details from the help file for the `prepare_diagram` function, which you can call like this

```{r, eval = FALSE}
help('prepare_diagram')
```

We'll explain the object returned from `prepare_diagram` next.

THIS SHOULD GO INTO HELP FILE.

Five (5) data frames are produced based on the model structure:

EXPLAIN WHICH NEED TO BE PRESENT AND WHICH CAN BE MISSING/EMPTY

1. `nodes`: a data frame with four columns - ALSO ROW?
    * `id`: the node (variable) id number, used to match with the edge data frames
    * `label`: the label of the node (variable), e.g. "S" for susceptible
    * `x`: the x position of the node box CENTER OR CORNER?
    * `y`: the y position of the node box
1. `horizontal_edges`: a data frame with 9 columns
    * `to`: the `id` of the node *to which* the flow terminates (e.g., arrow points toward this node)
    * `from`: the `id` of the node *from which* the flow origins (e.g., arrow points away this node)
    * `label`: the label of the flow to appear next to the line segment, typically an equation
    * `xstart`: the starting x position of the line segment
    * `ystart`: the starting y position of the line segment
    * `xend`: the ending x position of the line segment
    * `yend`: the ending y position of the line segment
    * `xmid`: the x midpoint of the line segment, for intersections with other segments due to interactions
    * `ymid`: the y midpoint of the line segment, for intersections with other segments due to interactions
1. `vertical_edges`: a data frame with 9 columns (same as `horizontal_edges`)
1. `curved_edges`: a data frame with 9 columns (same as `horizontal_edges`)
1. `feedback_edges`: a data frame with 9 columns (same as `horizontal_edges`)

The package attempts to make intelligent choices on which flows belong in which data frame.
For example, flows into or out of the system with only one node intersection are typically vertical.
And flows that bypass a node are typically curved.
The data frames above are returned as a list.
In some cases, many of the data frames will be empty.


# Modifying the `make_diagram` input object 

Once you understand the components of the input object and what the different parts mean, you can modify them.
This allows you to impact the look of the diagram you get once you send your modified object into the `make_diagram` function.

We'll show a few examples next.


# Example 1

Let's revisit the SIR model from the [basic modification vignette](./B_modify_diagrams_simple.html). 
First we set up the model.

```{r}
#Set up model
varlabels = c("S","I","R")
varnames = c("Susceptible","Infected","Recovered")
# currently only works with trailing + signs
flows = list(S_flows = c("+n", "-b*S*I", "-m*S"), 
             I_flows = c("+b*S*I","-g*I", "-m*I"), 
             R_flows = c("+g*I", "-m*R"))
sirmodel = list(varlabels = varlabels, varnames = varnames, flows = flows)
```

Next, we send the model to the `prepare_diagram` function.

```{r}
sir_input_structure <- prepare_diagram(sirmodel)
```

If we send this input structure to `make_diagram`, we get the following plot
```{r}
plot1 <- make_diagram(sir_input_structure)
plot(plot1)
```

Next, instead of sending this structure to the `make_diagram` function right away, we'll take a look at it and modify it.
Here are the details contained in this object

```{r}
print(sir_input_structure)
```

Let's modify this object somewhat. The following code moves one of the arrows and ...

```{r}
input_structure_modified <- sir_input_structure
input_structure_modified$curved_edges$xend[1] = 5.5
input_structure_modified$nodes$x[3] = 13 
```

With those modifications, the resulting diagram looks like this

```{r}
plot2 <- make_diagram(input_structure_modified)
plot(plot2)
```





# Example 2

Here is another example, the predator prey model you saw in earlier vignettes.

