---
title: Modifying the input structure 
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Modifying the input structure}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7
)

# avoid check where vignette filename must equal the vignette title
options(rmarkdown.html_vignette.check_title = FALSE)
```


```{r setup, include = FALSE}
pkg = 'flowdiagramr' #so we can easily switch names
library(pkg, character.only = TRUE)
```

## Introduction

We assume you went through the [getting started vignette](./A_quickstart.html) and also the [basic modification vignette](./B_modify_diagrams_simple.html). 

If the modifications you can achieve by providing optional settings to `prepare_diagram` and `make_diagram`, then you do not need to go any further, and there is no need to understand in more detail how **`r pkg`** works. However, it is quite likely that at some point, you want to further tweak your diagram. To do so successfully, it is useful to learn some more details about how the package works. 

As you learned, producing a diagram with **`r pkg`** involves two steps. After you have defined your model, there is a first step in which you call `prepare_diagram`, which takes your model and creates a list of data frames (which we called the `input_structure` in the prior vignettes). These data frames are used by `make_diagram` to produce the ggplot code and figure. 

The advantage of this 2-step process is that you can make modifications to the `input_structure` object before sending it into the `make_diagram` function, and thus modify the layout of the diagram beyond what you can through the`make_diagram` options, while still being fairly easy to do. 




## Explanation of the `diagram_list` object

To modify the object returned from `prepare_diagram` and used as input for `make_diagram`, it is important to learn a bit about it.
You can learn the details from the help file for the `prepare_diagram` function, which you can call like this

```{r, eval = FALSE}
help('prepare_diagram')
```

As the help file explains, what `prepare_diagram` does is to produce a list (which we call `diagram_list`, but you can assign it any name) that contains 2 data frames, one with variable related information and one with flow related information. This information is used by `make_diagram` to create the plot. 

`prepare_diagram` attempts to make reasonable choices on location and sizing for variables and flows. For example, flows into or out of the system with only one node intersection are typically drawn at a 45 degree angle. Flows that bypass a node are typically curved.
 
Once you understand the components of the input object and what the different parts mean, you can modify them.
This allows you to impact the look of the diagram you get once you send your modified object into the `make_diagram` function.

It is easiest to understand this approach by showing a few examples, so let's do that.


## Modifying the `make_diagram` input object 

### Example 1

Let's revisit the SIR model from the [basic modification vignette](./B_modify_diagrams_simple.html). 
First we set up the model.

```{r}
#Set up model
varlabels = c("S","I","R")
flows = list(S_flows = c("n", "-b*S*I", "-m*S"), 
             I_flows = c("b*S*I","-g*I", "-m*I"), 
             R_flows = c("g*I", "-m*R"))
sirmodel = list(varlabels = varlabels, flows = flows)
```

Next, we send the model to the `prepare_diagram` function.

```{r}
sir_diagram_list <- prepare_diagram(sirmodel)
```

Now, let's make and plot the diagram. For this plot, we also turned on the grid. This is useful in figuring out by how much to shift things around.

```{r}
plot1 <- make_diagram(sir_diagram_list, diagram_settings = list(with_grid = TRUE))
plot(plot1)
```

So far, nothing really new, apart from showing the grid in the plot.

Now, instead of sending the object produced by `prepare_diagram` (here called `sir_diagram_list`) to the `make_diagram` function right away, we'll take a look at it and modify it.
Here are the details contained in this object

```{r}
print(sir_diagram_list)
```

Let's modify this object somewhat. The following code makes the `I` box bigger, and moves the inflow and outflow arrows such that they start/end on the box with the new size. It also shifts the curved arrow and adjusts the label positioning.

```{r}
#assign to a new variable so we preserve the original
modified_diagram_list <- sir_diagram_list 
#increase the I box a bit both on left and right 
modified_diagram_list$variables$xmin[2] = 5
modified_diagram_list$variables$xmax[2] = 7
#modify inflow and outflow arrows to align with new box size 
modified_diagram_list$flows$xend[1] = 5
modified_diagram_list$flows$xstart[2] = 7
#move the b*S*I arrow end a bit
modified_diagram_list$flows$xend[7] = 4
#move the b*S*I label 
modified_diagram_list$flows$labely[7] = 0.2
modified_diagram_list$flows$labelx[7] = 4.6
#move the g*I arrow over just a bit
modified_diagram_list$flows$labelx[2] = 7.75
```

With those modifications, the resulting diagram looks like this

```{r}
plot2 <- make_diagram(modified_diagram_list)
plot(plot2)
```


It is possible to combine the modification of the diagram with modifications made through settings in `make_diagram`, e.g., we can add some color like this

```{r}
plot3 <- make_diagram(
  modified_diagram_list,
  diagram_settings = list(var_fill_color = c("#6aa4c8", "#eb5600", "#1a9988"),
                          var_label_color = "black",
                          main_flow_label_color = "blue",
                          interaction_flow_color = "red")
)
plot(plot3)
```



### Example 2

Here is another example, the predator prey model you saw in earlier vignettes.

Setting up the model:

```{r}
varlabels = c("Pat","Imm")
flows     = list(Pat_flows = c("g*Pat*(1-Pat/pmax)", "-dP*Pat", "-k*Pat*Imm"),
                 Imm_flows = c("r*Pat*Imm", "-dI*Imm"))
mymodel = list(varlabels = varlabels, flows = flows)
```

Preparing the diagram and making the plot.

```{r}
diagram_list <- prepare_diagram(mymodel)
ppplot <- make_diagram(diagram_list)
plot(ppplot)
```

Let's modify this. We'll make the `Imm` box larger, relabel it, adjust the location of some of the flows to work with the different box sizes. 
We'll also move some flow labels around and let's also assume that you want the parameter that specifies the maximum pathogen level at which growth goes to zero to be called `Pmax`, not `pmax`. You can't do that when you specify the model since **`r pkg`** requires that parameters start with a lower case letter. Not a big deal, you can fix this now by replacing the label.

First, we'll look at the `diagram_list` structure to see which elements we need to change.
```{r}
print(diagram_list)
```

And here is the code that makes some changes

```{r}
#increase box
diagram_list$variables$xmin[2] = 2.25
diagram_list$variables$xmax[2] = 3.75
#relabel box
diagram_list$variables$plot_label[2] = "T cells"
#adjust arrow location
diagram_list$flows$xstart[5] <- 2.25
#move labels
diagram_list$flows$labelx[4] <- 2
diagram_list$flows$labelx[5] <- 1.4
#change label
diagram_list$flows$label[6] <- "g*Pat*(1-Pat/Pmax)"
```

Sure enough, if you plot this now, you can see all your improvements.

```{r}
my_diagram <- make_diagram(diagram_list)
plot(my_diagram)
```

### Example 3

Finally, let's briefly revisit our 7-compartment model and see if we can make some further tweaks to get it to something that might be suitable for public use.

Here is the model specification again

```{r}
varlabels = c("Sc","Ic","Rc","Sa","Ia","Ra","P")
flows = list(Sc_flows = c("-Sc*bcc*Ic","-Sc*bca*Ia","-Sc*bcp*P"),
             Ic_flows = c("Sc*bcc*Ic","Sc*bca*Ia","Sc*bcp*P","-gc*Ic"),
             Rc_flows = c("gc*Ic"),
             Sa_flows = c("-Sa*bac*Ic","-Sa*baa*Ia","-Sa*bap*P"),
             Ia_flows = c("Sa*bac*Ic","Sa*baa*Ia","Sa*bap*P","-ga*Ia"),
             Ra_flows = c("ga*Ia"),
             P_flows = c("sc*Ic","sa*Ia","-d*P")
             )
mymodel = list(varlabels = varlabels, flows = flows)

varnames = c("Susceptible children","Infected children","Recovered children","Susceptible adults","Infected adults","Recovered adults","Pathogen in Environment")
varlocations = matrix(data = c("Sc", "Ic", "Rc",
                               "",   "P",   "",
                               "Sa", "Ia", "Ra"),nrow = 3, byrow = TRUE)
mysettings = list(varlocations = varlocations, varnames = varnames, use_varnames = TRUE)
diagram_list <- prepare_diagram(mymodel, mysettings)
```

If you print the resulting `diagram_list`, you can see details for each box/variable and arrow/flow. Since this is big, we are not showing it here. But we'll go ahead and make some modifications.

We'll increase the box for the pathogen and change the text for all variables by including both the full name and the label (this could of course also be done by just renaming `varnames` above, but we want to show a different way). We'll also move some of the flow labels so they don't overlap with the arrows.

**MODIFICATIONS NOT YET IMPLEMENTED, WILL BE DOING THAT ONCE ARROW LOCATION AND BOX SIZING IS ADDRESSED.**

```{r}
# change labeling of variables
#diagram_list$nodes$name = paste0(diagram_list$nodes$name,", ",diagram_list$nodes$label)
#increase box
#diagram_list$nodes$xmin[7] = 2.25
#diagram_list$nodes$xmax[7] = 3.75
#adjust arrow location
#diagram_list$curved_edges$xstart[2] <- 2.25
#move labels
#diagram_list$curved_edges$labelx[1] <- 2
#diagram_list$curved_edges$labelx[2] <- 1.4
```


Finally, we make the diagram, and also apply different styling as previously

```{r}
diagram_settings <- list(
                        var_fill_color = c("#6aa4c8", "#eb5600", "#1a9988", "#2987c2", "#e38b59", "#5c948c", "#e8e656"),
                        var_outline_color = "black"
                        )
model_plot <- make_diagram(diagram_list, diagram_settings)
plot(model_plot)
```




## Next steps

You now learned two ways that allow you to customize the diagram in a fairly easy way. The first approach is [through optional arguments to `make_diagram`](./B_modify_diagrams_simple.html), and the second approach is the one we just discussed. 

We expect that this is often enough to give you a diagram that's good enough for what you want. However, at times you might want to make further changes. For instance you might want to change specific elements, add text to the diagram, or perform other fine tuning. **`r pkg`** gives you a way of doing this, which is described in the [modifying the ggplot2 code vignette](./D_modify_ggplot_code.html), which you should read next. 


