---
title: Modifying the input structure 
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Modifying the input structure}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

# avoid check where vignette filename must equal the vignette title
options(rmarkdown.html_vignette.check_title = FALSE)
```


```{r setup, include = FALSE}
pkg = 'flowdiagramr' #so we can easily switch names
library(pkg, character.only = TRUE)
```

# Introduction

We assume you went through the [getting started vignette](./A_quickstart.html) and also the [basic modification vignette](./B_modify_diagrams_simple.html). 

If the basic modifications you can achieve by providing different options to `make_diagram`, then you do not need to go any further, and there is no need to understand in more detail how **`r pkg`** works. However, it is quite likely that at some point, you want to further tweak your diagram. To do so successfully, it is useful to learn some more details about how the package works. 

As you learned, producing a diagram with **`r pkg`** involves two steps. After you have defined your model, there is a first step that takes your model and creates a list of data frames (which we called the `input_structure` in the prior vignettes). These data frames are used by `make_diagram` to produce the ggplot code and figure. 

The advantage of this 2-step process is that you can make modifications to the `input_structure` object before sending it into the `make_diagram` function, and thus modify the layout of the diagram beyond what you can through the`make_diagram` options, while still being fairly easy to do. 




# Explanation of the `diagram_list` object

To modify the object returned from `prepare_diagram` and used as input for `make_diagram`, it is important to learn a bit about it.
You can learn the details from the help file for the `prepare_diagram` function, which you can call like this

```{r, eval = FALSE}
help('prepare_diagram')
```

As the help file explains, what `prepare_diagram` does is to produce a list (which we call `diagram_list`, but can assign it any name) that contains 5 elements, one for the node/variable placement information, and 4 others with information for different types of edges. Depending on the model, some of the edge data frames might be empty. This information about the location of the nodes/variables, and the different types of edges is used by `make_diagram` to create the plot. 

`prepare_diagram` attempts to make reasonable choices on node location and classification and location of flows. For example, flows into or out of the system with only one node intersection are typically drawn at a 45 degree angle. Flows that bypass a node are typically curved.
 
Once you understand the components of the input object and what the different parts mean, you can modify them.
This allows you to impact the look of the diagram you get once you send your modified object into the `make_diagram` function.

It is easiest to understand this approach by showing a few examples, so this is what we'll do next.


# Modifying the `make_diagram` input object 

## Example 1

Let's revisit the SIR model from the [basic modification vignette](./B_modify_diagrams_simple.html). 
First we set up the model.

```{r}
#Set up model
varlabels = c("S","I","R")
varnames = c("Susceptible","Infected","Recovered")
# currently only works with trailing + signs
flows = list(S_flows = c("+n", "-b*S*I", "-m*S"), 
             I_flows = c("+b*S*I","-g*I", "-m*I"), 
             R_flows = c("+g*I", "-m*R"))
sirmodel = list(varlabels = varlabels, varnames = varnames, flows = flows)
```

Next, we send the model to the `prepare_diagram` function.

```{r}
sir_diagram_list <- prepare_diagram(sirmodel)
```

If we send this input structure to `make_diagram`, we get the following plot
```{r}
#plot1 <- make_diagram(sir_diagram_list, with_grid = TRUE)
plot1 <- make_diagram(sir_diagram_list)
plot(plot1)
```

**NOT WORKING** For this plot, we also turned on the grid. This is useful in figuring out by how much to shift things around.


Next, instead of sending this structure to the `make_diagram` function right away, we'll take a look at it and modify it.
Here are the details contained in this object

```{r}
print(sir_diagram_list)
```

Let's modify this object somewhat. The following code makes the `I` box bigger, and moves the inflow and outflow arrows such that they start/end on the box with the new size. It also shifts the curved arrow and adjusts the label positioning.

```{r}
#assign to a new variable so we preserve the original
modified_diagram_list <- sir_diagram_list 
#increase the I box a bit both on left and right 
modified_diagram_list$nodes$xmin[2] = 8
modified_diagram_list$nodes$xmax[2] = 10
#modify inflow and outflow arrows to align with new box size 
modified_diagram_list$horizontal_edges$xend[1] = 8
modified_diagram_list$horizontal_edges$xstart[2] = 10
#move the b*S*I arrow end a bit
modified_diagram_list$curved_edges$xend[1] = 7
#move the b*S*I label up
modified_diagram_list$curved_edges$labely[1] = -1
```

With those modifications, the resulting diagram looks like this

```{r}
plot2 <- make_diagram(modified_diagram_list)
plot(plot2)
```


It is possible to combine the modification of the diagram with modifications made through settings in `make_diagram`, e.g., we can add some color like this

```{r}
plot3 <- make_diagram(modified_diagram_list,
                      diagram_settings = list(flow_text_color = "blue",
                                              node_fill_color = c("#6aa4c8", "#eb5600", "#1a9988"),
                                              node_text_color = "white"
                                              )
                     )
plot(plot3)
```



# Example 2

Here is another example, the predator prey model you saw in earlier vignettes.

Setting up the model:

```{r}
varlabels = c("Pat","Imm")
flows     = list(Pat_flows = c("g*Pat*(1-Pat/pmax)", "-dP*Pat", "-k*Pat*Imm"),
                 Imm_flows = c("r*Pat*Imm", "-dI*Imm"))
mymodel = list(varlabels = varlabels, flows = flows)
```

Preparing the diagram and making the plot.

```{r}
diagram_list <- prepare_diagram(mymodel)
ppplot <- make_diagram(diagram_list)
plot(ppplot)
```

There are a few things we might want to adjust. Let's make the `Imm` box larger, adjust the location of some of the flow labels and let's also assume that you want the parameter that specifies the maximum pathogen level at which growth goes to zero to be called `Pmax`, not `pmax`. You can't do that when you specify the model since `r pkg` requires that parameters start with a lower case letter. Not a big deal, you can fix this now by replacing the label.

First, we'll look at the `diagram_list` structure to see which elements we need to change.
```{r}
print(diagram_list)
```

**Need to be able to change label location for all flows. Not currently working.**

And here is the code that makes some changes

```{r}
#increase box
diagram_list$nodes$xmin[2] = 5.25
diagram_list$nodes$xmax[2] = 6.75
#adjust arrow location
diagram_list$curved_edges$xstart[2] <- 5.25
#move labels
#WANT TO MOVE LOCATION OF OTHER LABELS ASSOCIATED WITH FLOWS, e.g. g*Pat*(1-Pat/pmax) LOCATION
#change label
diagram_list$feedback_edges$label[1] <- "g*Pat*(1-Pat/Pmax)"
```

Sure enough, if you plot this now, you can see all your improvements.

```{r}
my_diagram <- make_diagram(diagram_list)
plot(my_diagram)
```
