---
title: Technical details 
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Technical details}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

# avoid check where vignette filename must equal the vignette title
options(rmarkdown.html_vignette.check_title = FALSE)
```

```{r setup}
pkg = 'flowdiagramr' #so we can easily switch names
library(pkg, character.only = TRUE)
```

# Introduction

We assume you went through the [getting started vignette]() and also the [basic modification vignette](). 
If the basic modifications you can do give you what you want, there is no need to dig deeper and understand in more detail how **`r pkg`** works. However, it is quite likely that at some point, you want to further tweak your diagram. To do so successfully, it is useful to learn some more details about how the package works. 


As you learned, producing a diagram with **`r pkg`** involves two steps. After you have defined your model, there is a first step that takes your model and creates a list of data frames (which we called the `input_structure` in the prior vignettes). These data frames are used by `make_diagram` to produce the ggplot code and figure. Going back to the SIR model, these are the steps

```{r}
#Set up model
varlabels = c("S","I","R")
varnames = c("Susceptible","Infected","Recovered")
# currently only works with trailing + signs
flows = list(S_flows = c("+n", "-b*S*I", "-m*S"), 
             I_flows = c("+b*S*I","-g*I", "-m*I"), 
             R_flows = c("+g*I", "-m*R"))
sirmodel = list(varlabels = varlabels, varnames = varnames, flows = flows)
# step 1: prepare inputs 
sir_input_structure <- prepare_diagram(sirmodel)
# step 2: make diagram
sir_diagram <- make_diagram(sir_input_structure)
```

The object returned after calling `prepare_diagram` contains data frames that have information stored in a way that it is fairly easy to use in ggplot diagram generation.

Here are the details contained in this object

Five (5) data frames are produced based on the model structure:

EXPLAIN WHICH NEED TO BE PRESENT AND WHICH CAN BE MISSING/EMPTY

1. `nodes`: a data frame with four columns - ALSO ROW?
    * `id`: the node (variable) id number, used to match with the edge data frames
    * `label`: the label of the node (variable), e.g. "S" for susceptible
    * `x`: the x position of the node box CENTER OR CORNER?
    * `y`: the y position of the node box
1. `horizontal_edges`: a data frame with 9 columns
    * `to`: the `id` of the node *to which* the flow terminates (e.g., arrow points toward this node)
    * `from`: the `id` of the node *from which* the flow origins (e.g., arrow points away this node)
    * `label`: the label of the flow to appear next to the line segment, typically an equation
    * `xstart`: the starting x position of the line segment
    * `ystart`: the starting y position of the line segment
    * `xend`: the ending x position of the line segment
    * `yend`: the ending y position of the line segment
    * `xmid`: the x midpoint of the line segment, for intersections with other segments due to interactions
    * `ymid`: the y midpoint of the line segment, for intersections with other segments due to interactions
1. `vertical_edges`: a data frame with 9 columns (same as `horizontal_edges`)
1. `curved_edges`: a data frame with 9 columns (same as `horizontal_edges`)
1. `feedback_edges`: a data frame with 9 columns (same as `horizontal_edges`)

The package attempts to make intelligent choices on which flows belong in which data frame.
For example, flows into or out of the system with only one node intersection are typically vertical.
And flows that bypass a node are typically curved.
The user may need to make modifications to suite their particular needs and style.

The data frames above are returned as a list.
In some cases, many of the data frames will be empty, as is this case in this example with a relatively simple model.

Here is an example of the list of data frames for the SIR model.

```{r}
print(sir_input_structure)
```




Here is another example, the predator prey model you saw in earlier vignettes.

```{r}
varlabels = c("B","I")
varnames = c("Bacteria","Immunity")
flows = list(B_flows = c("g*B", "-dB*B", "-k*B*I"), 
             I_flows = c("r*B*I", "-dI*I"))
#this is the actual model that needs to be used
#but currently it doesn't work, prepare_diagram complains
#flows = list(B_flows = c("g*B*(1-B/Bmax)", "-dB*B", "-k*B*I"), 
#                I_flows = c("r*B*I", "-dI*I"))
ppmodel = list(varlabels = varlabels, varnames = varnames, flows = flows)
pp_input_structure <- prepare_diagram(ppmodel)
print(pp_input_structure)
```

