---
title: Modifiying diagrams with `update_diagram()`  
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Modifiying diagrams with `update_diagram()`}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7
)

# avoid check where vignette filename must equal the vignette title
options(rmarkdown.html_vignette.check_title = FALSE)
```

```{r setup, include = FALSE}
pkg = 'flowdiagramr' #so we can easily switch names
library(pkg, character.only = TRUE)
```


## Introduction

We assume you went through the previous vignettes, especially the [getting started vignette](./A_quickstart.html). 
In this vignette, you will learn how to use the `update_diagrm()` function to further customize the look of your diagrams.

To re-cap, the workflow discussed in the previous vignettes goes like this

```{r}
#specify model
sirmodel1 = list(variables = c("S","I","R"),
                 flows     = list(S_flows = c("-b*S*I"), 
                                  I_flows = c("b*S*I","-g*I"), 
                                  R_flows = c("g*I")))
#optional layout information
sirsettings1 = list(varlocations = matrix(data = c("S", "","R",
                                                   "", "I","" ),
                                         nrow = 2, ncol = 3, byrow = TRUE),
                   varbox_x_size = c(1, 1.5, 1),
                   varbox_y_size = c(2, 1, 1),
                   varspace_x_size = c(1.5,2),
                   varspace_y_size = 0.5
                  )
#prepare diagram
diagram_list1 <- prepare_diagram(model_list = sirmodel1, model_settings = sirsettings1)
#make diagram
sir_diagram1 <- make_diagram(diagram_list = diagram_list1)
#plot diagram
plot(sir_diagram1)
```



## A quick look behind the scenes

Before we look at examples, let's briefly discuss some aspects of what's going on behind the scenes. 

The function `prepare_diagram()` takes the model and optional settings and returns a list (which we call `diagram_list` here) consisting of 2 data frames, one called `variables` and one called `flows`. In each data frame, the elements (variables or flows) are in the rows, and the columns indicate graphical settings, e.g. the color. You can look at the `diagram_list` data frames. Here are those two data frames for the model above:

```{r}
print(diagram_list1$variables)
```

```{r}
print(diagram_list1$flows)
```



When **`r pkg`** prepares the diagram, it groups flows into 3 different categories as follows:

* Main flows, which are flows that go from one box to another, and generally indicate physical/actual movement of units from one state to another (e.g., from susceptible to infected). 
* Interaction flows, which indicate a variable interacting with another, without any actual flow (e.g., infected individuals infecting susceptible).
* External flows, which indicate entry into or exit out of the system, e.g., births and deaths to/from states that are not part of the model. (The model above has no external flows.)

Each of those flow types receives its own default styling. You can change that as you like. It is also quite possible that sometimes **`r pkg`** doesn't guess correctly and labels a flow differently than what it should be. In that case, you can also easily overwrite the default settings.


In general, before you make the diagram with `make_diagram()`, you can adjust all the settings for each variable/box or flow/arrow. A direct way is to just edit the elements of the two data frames by hand. For instance this code changes the color of the box for the `I` variable to red and increases the size of the interaction flow showing the infection process.

```{r}
#make a copy
diagram_list2 <- diagram_list1
#change some styling
diagram_list2$variables$fill_color[2] <- "red"
diagram_list2$flows$line_size[3] <- 1.5
```

This is how the updated diagram looks.

```{r}
sir_diagram2 <- make_diagram(diagram_list2)
plot(sir_diagram2)
```

This way of manually updating is always possible, but it is a bit tedious. To make updating the look of the various plot elements easier, **`r pkg`** contains a  convenience function called `update_diagram()`, which simply does the updating of the elements in either of those data frames based on the supplied input. The function needs two inputs. The first input is the list returned by `prepare_diagram()`. The second input is a list of named entries which specify what elements of the diagram you want to update. The help file for `update_diagram()` provides all the details for allowed inputs. The examples we explore now show several of the options. 



## Example 1

We'll start by repeating the changes we made above, now using `update_diagram()`.

To update specific elements (variables or flows), you need to provide their names, as you will see below. You can get the names of all elements by calling `update_diagram()` with only the `diagram_list` element, like so:

```{r}
update_diagram(diagram_list1)
```

If you wanted to make the same changes as we did manually above, this would do it


```{r}
diagram_list3 <- update_diagram(diagram_list1, diagram_settings = list(var_fill_color = c(I = "red"),
                                                                       flow_line_size = c(i_bSI = 1.5) 
                                                                       ) 
                                )
```

This produces the same figure as above:

```{r}
sir_diagram3 <- make_diagram(diagram_list3)
plot(sir_diagram3)
```



When you modify any input with `update_diagram()`, you generally have two options. You can either specify individual elements you want to update, which we just did. For convenience, there is also a way to update multiple elements at the same time. Let's say we wanted to update the outline colors of all variables and the colors of all main flows, this would do it:


```{r}
#define updated settings
diagram_settings1 = list(var_fill_color = c(all = "green"),
                        flow_line_color = c(main = "orange")
                        )
#update settings
diagram_list4 <- update_diagram(diagram_list1, diagram_settings1)
sir_diagram4 <- make_diagram(diagram_list4)
plot(sir_diagram4)
```


The reserved words for `update_diagram()` are `all` (applied to variables and flows) and `main`, `interaction` and `external` (only applied to flows.)




## Example 2

Next, we are revisiting one of the models discussed in the introductory vignette, the predator-prey model for pathogen and immune response. Here is the model again.

```{r}
variables = c("Pat","Imm")
flows     = list(Pat_flows = c("g*Pat*(1-Pat/pmax)", "-dP*Pat", "-k*Pat*Imm"),
                 Imm_flows = c("r*Pat*Imm", "-dI*Imm"))
mymodel = list(variables, flows)
```

Here is the step that prepares the model, without any optional settings (for now).

```{r}
diagram_list_v1 <- prepare_diagram(mymodel)
```

This is how the default plot looks.

```{r}
diagram_v1 <- make_diagram(diagram_list_v1) 
plot(diagram_v1)
```

Using the `update_diagram()` function, it is possible to change the look of most components of the diagram. First, we determine the names of the different variables and flows:

```{r}
update_diagram(diagram_list_v1)
```
Note that the flow names are simply the equations without the math symbols, and indexed by type of flow. That can make them look a bit ugly at times, but it shouldn't be much of an issue.

Now we alter the look:

```{r}
# define new settings
diagram_settings2 = list(var_fill_color = c(Pat = "blue", Imm = "red"), 
                         var_label_color = c(all = "orange"),
                         flow_linetype = c(main = "dashed"),
                         flow_line_color = c(m_gPat1Patpmax = "green", m_ = "orange"),
                         flow_line_size = c(m_gPat1Patpmax = 1.5, e_dIImm = 2.5, interaction = 3.5),
                         flow_label_size = c(e_dIImm = 8, external = 2, interaction = 12),
                         flow_label_color = c(external = "cyan")
                        )
#update diagram list
diagram_list_v2 <- update_diagram(diagram_list_v1,diagram_settings2)
```

The result of those modifications is this fairly messy looking diagram:

```{r}
diagram_v2 <- make_diagram(diagram_list_v2) 
plot(diagram_v2)
```

Note that for some settings, we used a mix of addressing elements by name and some of the grouping shorthand. If you supply both, the way `update_diagram()` works is to first process the groupings (`all`/`main`/`interaction`/`external`) and then any individual ones. This means you can for instance change all boxes to one color and then individually set one to yet another color. We'll show that in the next example.


## Example 3

Let's revisit the SIR model with natural births and deaths

```{r}
# specify the model
variables = c("S","I","R")
flows = list(S_flows = c("n", "-b*S*I", "-m*S"), 
             I_flows = c("+b*S*I","-g*I", "-m*I"), 
             R_flows = c("g*I", "-m*R"))
sirmodel = list(variables = variables, flows = flows)
```


Here are the defaults

```{r}
#prepare diagram
diagram_list1 <- prepare_diagram(model_list = sirmodel)
#make diagram
sir_diagram1 <- make_diagram(diagram_list = diagram_list1)
#plot diagram
plot(sir_diagram1)
```


Here is an alternative version with layout changes specified through `update_diagram()`.

First, we get the names of all elements:

```{r}
update_diagram(diagram_list1)
```
Now we update the look of various elements. 

```{r}
#define new diagram settings outside the update function so we can reuse
  diagram_settings = list(
    var_fill_color = c(I = "#FFB3A8", all = "#b2b2b2"),
    var_outline_color = c(all = "#031e79"),
    var_label_color = c(all = "black"),
    flow_line_color = c(interaction = "#f94075", external = "blue"),
    flow_show_label = c(interaction = FALSE, e_n = FALSE),
    flow_show_arrow = c(e_n = FALSE)
  )
#update diagram
diagram_list2 <- update_diagram(diagram_list1,diagram_settings)
#make diagram
sir_diagram2 <- make_diagram(diagram_list = diagram_list2)
#plot diagram
plot(sir_diagram2)
```

Note again the mix of updates applied to individual elements and groups of elements. Groupings are always applied first, no matter the order you supply them. Thus, first we set the color of all boxes to gray, then change the `I` box individually to red.


## Example 4

It is of course entirely possible to combine layout settings supplied to `prepare_diagram()` with further styling through `update_diagram()`. Here is an example, using the SIR model we just explored.


```{r}
#optional layout settings
sirsettings = list(varlocations = matrix(data = c("S", "","R",
                                                   "", "I","" ),
                                         nrow = 2, ncol = 3, byrow = TRUE),
                   varbox_x_size = c(1, 1.5, 1),
                   varbox_y_size = c(1, 1, 1),
                   varspace_x_size = c(1,1),
                   varspace_y_size = 0.5
                  )

#prepare diagram
diagram_list3 <- prepare_diagram(model_list = sirmodel, model_settings = sirsettings)
#change styling
diagram_list4 <- update_diagram(diagram_list3,diagram_settings)
#make diagram
sir_diagram4 <- make_diagram(diagram_list = diagram_list4)
#plot diagram
plot(sir_diagram4)
```


## Example 4

Finally, we'll revisit the 7-compartment model we explored previously, and see if we can get closer to something that could be shown to the public `r emoji::emoji('smile')`.

```{r}
variables = c("Sc","Ic","Rc","Sa","Ia","Ra","P")
flows = list(Sc_flows = c("-Sc*bcc*Ic","-Sc*bca*Ia","-Sc*bcp*P"), 
             Ic_flows = c("Sc*bcc*Ic","Sc*bca*Ia","Sc*bcp*P","-gc*Ic"), 
             Rc_flows = c("gc*Ic"),
             Sa_flows = c("-Sa*bac*Ic","-Sa*baa*Ia","-Sa*bap*P"), 
             Ia_flows = c("Sa*bac*Ic","Sa*baa*Ia","Sa*bap*P","-ga*Ia"), 
             Ra_flows = c("ga*Ia"),
             P_flows = c("sc*Ic","sa*Ia","-d*P")
             )
mymodel = list(variables, flows)
```

```{r}
mysettings = list( varlocations = matrix(data = c("Sc", "", "Ic", "Rc",
                                                  "",   "P", "",   "",
                                                  "Sa", "", "Ia", "Ra"),nrow = 3, byrow = TRUE),
                   varspace_x_size = 2,
                   varspace_y_size = 1
                   )
```

Prepare the diagram with  additional settings.

```{r}
diagram_list5 <- prepare_diagram(mymodel, mysettings)
```

Update look of diagram

```{r}
#get element names
update_diagram(diagram_list5)
```


```{r}
# define list of updates
diagram_settings2 = list(var_fill_color = c(Sc = "#cde9fa", Ic = "#cde9fa", 
                                            Rc = "#cde9fa", Sa = "#40f9cf", 
                                            Ia = "#40f9cf", Ra = "#40f9cf", 
                                            P = "#b2b2b2"),
                         var_outline_color = c(all = "#031e79"),
                         var_label_color = c(all = "black"),
                         flow_line_color = c( interaction = "#f94075",
                                              external = "gray")
                         )
diagram_list6 <- update_diagram(diagram_list5, diagram_settings2) 
```


Now we make the diagram and plot it.

```{r}
model_plot6 <- make_diagram(diagram_list6)
plot(model_plot6)
```
Better, but not quite there. One can use `update_diagram()` more than once of course, so let's make further changes. Specifically, we'll modify some of the locations for the arrows and labels.

AH: WAITING TO DO THIS UNTIL WE FINALIZED HOW UPDATES TO LOCATIONS WORK, SEE NOTE IN TODO.

```{r}
# define list of updates
diagram_settings3 = list(flow_ylabel = c(m_gcIc = 0.5)
                         )
diagram_list7 <- update_diagram(diagram_list6, diagram_settings3) 
```



```{r}
model_plot7 <- make_diagram(diagram_list7)
plot(model_plot7)
```

This is now looking like a pretty decent diagram. Further tweaks are of course possible.



## Next steps

By combining layout settings supplied to `prepare_diagram()` and styling applied through `update_diagram()`, you can easily customize the look of the diagram. Often, this should allow you to produce a diagram that looks ready to use. 
In thos instances where you want to make further changes that are not currently supported by this workflow (e.g., if you want to change one of the variable boxes to a circle), **`r pkg`** can provide you with the complete code that generates the `ggplot` object, which you can then edit manually.

[The next vignette](./D_modify_input_structure.html) explains how to do that. 






