---
title: Modifying the ggplot2 object or code 
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Modifying the ggplot2 object or code}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: sentence
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7
)

# avoid check where vignette filename must equal the vignette title
options(rmarkdown.html_vignette.check_title = FALSE)
```

```{r setup, include = FALSE}
pkg = 'flowdiagramr' #so we can easily switch names
library(pkg, character.only = TRUE)
library(ggplot2)
```

## Introduction

We assume you went through the previous vignettes that showed you how to use the functionality provided by **`r pkg`** to produce fairly good looking (in our biased opinion) diagrams.
We believe (hope) that this often gives you a diagram you are happy with.

However, occasionally you might want to make additional tweaks beyond what **`r pkg`** can do.
Fortunately for you, there is a way to do this, which is the topic of this vignette.
There's one easy way, and one approach that's a bit more involved but also more flexible.

## Modifying the ggplot object

The object returned by `make_diagram()` is a ggplot object, which means you can do all the things you can do to those objects.
That includes further alterations and updates.
Here are a few simple examples.

### Example 1

Back to our simple SIR model

```{r}
# specify the model
variables = c("S","I","R")
flows = list(S_flows = c("n", "-b*S*I", "-m*S"), 
             I_flows = c("+b*S*I","-g*I", "-m*I"), 
             R_flows = c("g*I", "-m*R"))
sirmodel = list(variables = variables, flows = flows)
# prepare, make and plot
diag_list <- prepare_diagram(sirmodel)
diag1 <- make_diagram(diag_list)
plot(diag1)
```

Let's assume you want to give it a title and add some more text into the plot.
You can just use the regular `ggplot2` syntax.

```{r}
diag2 <- diag1 + ggtitle("My diagram") + geom_label(aes(x=2,y=2),label="My label")
plot(diag2)
```

### Example 2

This example shows the overall suggested workflow, namely first using all the functionality that **`r pkg`** provides, and then further modifying the plot with `ggplot2` commands.

For this example, we revisit our somewhat complex 7-compartment model.
This is the code we had previously to get a decent figure.

Model definition


```{r}
variables = c("Sc","Ic","Rc","Sa","Ia","Ra","P")
flows = list(Sc_flows = c("-bcc*Sc*Ic","-bca*Sc*Ia","-bcp*Sc*P"), 
             Ic_flows = c("bcc*Sc*Ic","bca*Sc*Ia","bcp*Sc*P","-gc*Ic"), 
             Rc_flows = c("gc*Ic"),
             Sa_flows = c("-bac*Sa*Ic","-baa*Sa*Ia","-bap*Sa*P"), 
             Ia_flows = c("bac*Sa*Ic","baa*Sa*Ia","bap*Sa*P","-ga*Ia"), 
             Ra_flows = c("ga*Ia"),
             P_flows = c("rc*Ic","ra*Ia","-d*P")
             )
mymodel = list(variables, flows)
```

Basic layout settings

```{r}
mysettings = list( varlocations = matrix(data = c("Sc", "", "Ic", "Rc",
                                                  "",   "P", "",   "",
                                                  "Sa", "", "Ia", "Ra"),nrow = 3, byrow = TRUE),
                   varbox_x_size = 2,
                   varspace_x_size = 1.5
                   )
```

List of all styling updates

```{r}
# define list of updates
diagram_settings = list(var_fill_color = c(Sc = "#cde9fa", Ic = "#cde9fa", 
                                            Rc = "#cde9fa", Sa = "#40f9cf", 
                                            Ia = "#40f9cf", Ra = "#40f9cf", 
                                            P = "#b2b2b2"),
                         var_outline_color = c(all = "#031e79"),
                         var_label_size = c(all = 4),
                         var_label_color = c(all = "black"),
                         flow_label_size = c(all = 4),
                         flow_curvature = c(i_baaSaIa = -0.5, i_bccScIc = 0.5, 
                                            i_bacSaIc = -0.3, i_bcaScIa = 0.3,
                                            g_raIa = 0, g_rcIc = 0, i_bcpScP = -0.5),
                         flow_xstart = c(i_bcpScP = -1, i_bapSaP = -1),
                         flow_ystart = c(i_bcpScP = -0.5, i_bapSaP = -0.5, i_bacSaIc = -1.1, 
                                         i_baaSaIa = -1.1, i_bccScIc = 0.1),
                         flow_line_color = c(i_baaSaIa = "#40f9cf", i_bacSaIc = "#cde9fa",
                                             i_bapSaP = "orange", i_bcpScP = "orange",
                                             i_bccScIc = "#cde9fa", i_bcaScIa = "#40f9cf"),
                         flow_line_size = c(all = 1.1),
                         flow_xlabel = c(i_bcaScIa = 1, i_bacSaIc = 4, 
                                         i_bapSaP = -1.25, i_bcpScP = -2.25,
                                         g_raIa = 0.15, g_rcIc = 0.15), 
                         flow_ylabel = c(m_gcIc = 0.25, m_gaIa = 0.25,
                                         i_bcaScIa = -2.5, i_bacSaIc = -0.5, 
                                         i_baaSaIa = -2.15, i_bccScIc = -0.5, 
                                         e_dP = -0.5,
                                         g_raIa = 0.5, g_rcIc = -0.75)
                        
                         )
```

Prepare, update and make the diagram.

```{r}
diagram_list <- prepare_diagram(mymodel, mysettings)
diagram_list2 <- update_diagram(diagram_list, diagram_settings)
diag <- make_diagram(diagram_list2)
plot(diag)
```

Notice that we made a change above, we didn't add the full text to the boxes but kept the variable names. Let's assume we want to add the full text next to the boxes. We can simply add them as text to the `ggplot` object, like this:

```{r}
var_label_text = c(Sc = "Susceptible\n Children", Ic = "Infected\n Children", Rc = "Recovered\n Children",
                   Sa = "Susceptible\n Adults",   Ia = "Infected\n Adults",   Ra = "Recovered\n Adults",
                   P = "Pathogen in\n Environment")
xpos = c(2,9,12.5,2,9,12.5,3.5)
ypos = c(3.5,5.5,3.5, 1.5,-0.5,1.5, 2.5)
diag2 <- diag + ggtitle("7 compartment model") + 
                annotate("text", x = xpos, y = ypos, label = var_label_text)
plot(diag2)
```
This is not perfect and needs a bit more tweaking, but you get the idea. This approach allows you to add or modify the ggplot object returned from `r pkg` any way you want.


## Modifying the ggplot2 code

While the above approach toward modifying the `ggplot` object returned by `make_diagram()` gives you some additional flexibility, you can go a step further.
The most flexible approach to adjust diagrams is to modify the ggplot code that produces the diagram.
**`r pkg`** provides a function that returns the code, which you can then edit like you would any `ggplot` code.
We'll illustrate this approach with a few examples.

### Example 1

First, you define your model as usual.

```{r}
# specify the model
variables = c("S","I","R")
flows = list(S_flows = c("n", "-b*S*I", "-m*S"), 
             I_flows = c("+b*S*I","-g*I", "-m*I"), 
             R_flows = c("g*I", "-m*R"))
sirmodel = list(variables = variables, flows = flows)
```

Next, you prepare the model

```{r}
diag_list <- prepare_diagram(sirmodel)
```

Once you have the diagram list ready, you can send it to `write_diagram`, a function that writes the full code needed to generate the diagram.
You can specify the location and name for the file.
If you don't provide those, it will be stored in the current working directory under the name `diagram_code.R`.

```{r, eval= FALSE}
# write the current diagram code to a file, overwrite if it exists
write_diagram(diagram_list = diag_list, filename = 'model1_diagram_code.R', always_overwrite = TRUE)
```

If things worked well, you should get a message telling you the location where the code was saved.
The code that is generated is completely self-contained, that means you can give it to anyone, who can run it and reproduce your model figure.

If you run the code that is generated, it will create the diagram structure as an object called `diagram_plot`.
You can then plot or save that object (there is code at the bottom of the generated file for that).

For this example, here is what you get.

```{r, eval = TRUE}
source('model1_diagram_code.R')
plot(diagram_plot)
```

Having all the code that generates the figure allows you to do some fine-tuning.
We provided hopefully enough comments in the code to make it clear what happens where.
That said, to be able to tweak the diagram this way requires being familiar with creating figures in ggplot.

Let's assume for this example that you want to add some text below the `I` compartment.
In this case, you don't need to modify any of the existing code, just add a some code at the end.
As shown above, you could do this by just modifying the `ggplot` object returned from `r pkg`. But here we'll do it by editing the code.

The following lines of code appended to the bottom of `sirmodel1_diagram_code.R`, just before the final `plot` statement, will do the trick:

```{r, eval=FALSE}
#######################
### LAST PART OF THE GENERATED CODE BEFORE ADDITION
#######################
with_grid <- FALSE  # default is false
if(with_grid == FALSE) {
  diagram_plot <- diagram_plot +
    theme_void()  # makes an empty plot theme with no axes, grids, or ticks
} else {
  # The else here may seem silly, but otherwise the returned plot is NULL
  diagram_plot <- diagram_plot  # just returns default ggplot2 theme
}

#######################
### YOUR NEW ADDITION
#######################
# make a new data frame of text
text_df <- data.frame(  x = 3,  y = -1,
  lab = "Infected and Infectious"
)
diagram_plot <- diagram_plot +
  geom_text(data = text_df, aes(x = x, y = y, label = lab), size = 8)

#######################
### PART OF THE GENERATED CODE AFTER ADDITION
#######################
# These lines plot or save the generated diagram. 
# Uncomment them if you want to perform either action. 
# plot(diagram_plot) 
# ggsave('diagram_plot.png',diagram_plot)
```

Then save the new code, let's call it `sirmodel1_diagram_code_modified.R`.
Running the new code will add the wanted text to the diagram plot

```{r, eval = TRUE}
source('model1_diagram_code_modified.R')
plot(diagram_plot)
```

### Example 2

Let's look at another example, one that requires this approach.
For that, we again consider the 7-compartment model.
Assume we done all our styling using the **`r pkg`** functions as above.
But we want to do something that is currently not supported, namely changing the shape of one of the variables.

To do that, we take the diagram list we produced above and write the code to a file, then edit it.

```{r, eval= FALSE}
write_diagram(diagram_list = diagram_list2,
              filename = 'model2_diagram_code.R',
              always_overwrite = TRUE)
```

If you run the code that is generated, you should see a figure like this

```{r, eval = TRUE}
source('model2_diagram_code.R')
plot(diagram_plot)
```

Now let's turn the shape of the `P` box into a circle.
You can find the part of the code that creates the boxes (it's the first for loop) and modify as follows:

```{r, eval = FALSE}
#######################
### LAST PART OF THE GENERATED CODE BEFORE ADDITION
#######################
for(i in 1:nrow(variables)) {

#######################
### YOUR NEW ADDITION
#######################
if(variables$name[i] == "P") {
    diagram_plot <- diagram_plot +
      geom_point(
        data = variables[i, ],
        aes(x = xmin+(xmax-xmin)/2, y = ymin+(ymax-ymin)/2), #centers
        color = variables[i, "outline_color"],
        fill = variables[i, "fill_color"],
        shape = 21,
        size = 23
      )
  } else {
  #######################
  ### END YOUR NEW ADDITION
  #######################
 diagram_plot <- diagram_plot +
    geom_rect(
      data = variables[i, ],
      aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
      color = variables[i, "outline_color"],
      fill = variables[i, "fill_color"]
       )
  }
}
```

Now save the modified script, source it and plot the generated `diagram_plot` object to get this figure:

```{r, eval = TRUE}
source('model2_diagram_code_modified.R')
plot(diagram_plot)
```

Again, not quite there yet. We would need to update placement of some of the arrows. Since we modified the ggplot code, we either need to make further modifications also in the ggplot code. Alternatively, one can adjust as much as possible previously. E.g., in this case, adjust arrow placement above in anticipation of this change later in the code.


## General suggestions

It should be clear from these examples that once you have the ggplot code, you can do any modifications you want.
At this point, the options are endless and you can change the code to get exactly what you want - at the expense of having to write/edit code yourself.

As mentioned before, the best workflow is to use the **`r pkg`** functionality to get as far as it takes you, and then as needed make further manual modifications.

Of course, once you have started manual editing of the code, things are not reversible, i.e. you have basically left the **`r pkg`** package infrastructure and are now adjusting the diagram on your own.
You can of course always re-generate the code and copy your modifications over (as long as you remember to not overwrite your code.)

## Next steps

This concludes the main tutorials for the package.
Several additional vignettes are available.
You can access all [from the package website](https://andreashandel.github.io/flowdiagramr/).

While **`r pkg`** was designed with compartmental/systems modeling in mind, you can use it to create diagrams for anything consisting of nodes/boxes and flows/arrows.
One of the vignettes provides some examples.

A fairly long vignette shows a lot of further model examples, and provides some tips on how to make good-looking diagrams even if **`r pkg`** doesn't fully cooperate.
We recommend you check that one out.

Another vignette discusses additional topics such as using **`r pkg`** together with our [modelbuilder package](https://ahgroup.github.io/modelbuilder/) and how to crop white space around the diagrams.

We also have a short vignette that compares **`r pkg`** to similar R packages.
