---
title: Using modelbuilder models with flowdiagramr
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using modelbuilder models with flowdiagramr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

# avoid check where vignette filename must equal the vignette title
options(rmarkdown.html_vignette.check_title = FALSE)
```

```{r setup, include = FALSE}
pkg = 'flowdiagramr' #so we can easily switch names
library(pkg, character.only = TRUE)
```


# Overview

We have an R package called [**modelbuilder**]() which allows users to graphically build .
The **`r pkg`** package is in some sense a spin-off from **modelbuilder** and is used as part of **modelbuilder**. As such, it is important that one can use **modelbuilder** built models with **`r pkg`**.

Since **modelbuilder** models and **`r pkg`** models are slightly different, we provide a convience function that converts one format into the other. This is a very simple process, as the following example shows.



# Example

Assume you built a model with **modelbuilder** and saved it as an `R` object inside a `.Rda` file. 


