---
title: More model examples
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{More model examples}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7
)

# avoid check where vignette filename must equal the vignette title
options(rmarkdown.html_vignette.check_title = FALSE)
```

```{r setup, include = FALSE}
pkg = 'flowdiagramr' #so we can easily switch names
library(pkg, character.only = TRUE)
```

# Overview

The previous tutorials/vignettes showed several example models as way to illustrate how to use **`r pkg`**.

This vignette is a (growing) list of further model examples, which might be helpful to illustrate further ways of using the package.


# Example 1

This is a model that describes the infection or vaccination dynamics of an acute viral infection in a simplified manner.
The model tracks virus/antigen (H), and a few immune response components. These are the differential equations for the model.

$$
\begin{aligned}
\textrm{Antigen} \qquad  \dot{H} & =   - kAH - c H \\ 
\textrm{Interferon} \qquad \dot{F} & = p - m F +  q (F_{max} - F) \frac{H}{H + n}   \\
\textrm{CD4 T-cells} \qquad \dot{T} & =  \frac{FHT}{FH + h_T} + g_TT   \\
\textrm{B-cells} \qquad \dot{B} & =  g B(\frac{r_H HF }{s_1 + HF} + \frac{r_T TF}{s_2 + TF})    \\
\textrm{Antibodies} \qquad \dot{A} & =   r B  - kAH - d A   \\
\end{aligned}
$$

This is the same information formatted as a structure that can be used as input for `prepare_diagram`.

```{r, error = TRUE}
varlabels = c("H","F","T","B","A")
varnames = c("Antigen","Interferon","T-cells","B-cells","Antibody")
flows = list(H_flows = c("-kA*A*H", "- dH * H"),
             F_flows = c("pF" ,"- dF*F", "H / (H + hH) * gF * (fmax - F)"),
             T_flows = c("F*H*T/(F*H + hT)" ,"+ gT*T"),
             B_flows = c("gB * B * rH*F*H/(s1+F*H)",  "gB * B *rT*F*T/(s2+F*T)" ),
             A_flows = c("rA*B",  "- dA*A", "- kA*A*H") 
             )
model_list = list(varlabels = varlabels, flows = flows, varnames = varnames)
diagram_list <- prepare_diagram(model_list)
```

# Example 2

This is the simple predator-prey model, with some styling applied.


First, we specify the model again. We did a bit of renaming here, what we called pathogen before is now called bacteria and labeled **B** and the Immune response is abbreviated as **I**. Parameter names have been adjusted accordingly, but otherwise it is the same model you already saw.

```{r}
# specify the model
varlabels = c("B","I")
flows = list(B_flows = c("+g*B*(1-B/bmax)", "-dB*B", "-k*B*I"),
             I_flows = c("+r*B*I", "-dI*I"))
ppmodel = list(varlabels = varlabels, flows = flows)
```

Next, we prepare the input list of data frames

```{r}
# prepare inputs 
pp_diagram_list <- prepare_diagram(ppmodel)
```

Finally, we can make a diagram with settings adjusted to our liking.

```{r}
# define adjusted settings for layout
diagram_settings = list(
    node_outline_color = NA,
    node_fill_color = c("#6aa4c8", "#eb5600", "#1a9988"),
    node_text_color = "white",
    node_text_size = 10,
    flow_text_color = "black",
    flow_text_size = 5,
    main_arrow_color = "grey25",
    main_arrow_linetype = 1,
    main_arrow_size = 0.7,
    interaction_arrow_color = "grey25",
    interaction_arrow_linetype = 2,
    interaction_arrow_size = 0.7
)
    
# make diagram with adjusted settings
diag1 <- make_diagram(
  diagram_list = pp_diagram_list,
  diagram_settings = diagram_settings
)

plot(diag1)
```

For the next diagram, we are turning off the interactions and only show physical flows, that means processes that actually move some entity (here bacteria and immune response) from one state to the next.

```{r}
# specify custom settings for the diagram layout
diagram_settings <- list(
    node_outline_color = NA,
    node_fill_color = c("#6aa4c8", "#eb5600", "#1a9988"),
    node_text_color = "white",
    node_text_size = 10,
    flow_text_color = "black",
    flow_text_size = 5,
    main_arrow_color = "grey25",
    main_arrow_linetype = 1,
    main_arrow_size = 0.7,
    interaction_label = FALSE,
    interaction_arrow_color = "grey25",
    interaction_arrow_linetype = 2,
    interaction_arrow_size = 0.7
)

    # make another diagram with adjusted settings
diag2 <- make_diagram(
  diagram_list = pp_diagram_list,
  diagram_settings = diagram_settings
)


plot(diag2)
```

As you can see, not showing the interactions doesn't produce a very useful diagram, so in this (and most cases), not showing interaction terms is maybe not the best idea.


            
