---
title: Comparison to other R packages
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comparison to other R packages}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

# avoid check where vignette filename must equal the vignette title
options(rmarkdown.html_vignette.check_title = FALSE)
```

```{r setup, include = FALSE}
pkg = 'flowdiagramr' #so we can easily switch names
library(pkg, character.only = TRUE)
```


# Overview

Here, we'll provide a brief discussion how **`r pkg`** differs from and can interact with other packages. 


# diagram

The **diagram** package (https://cran.r-project.org/web/packages/diagram/index.html) is most similar to **`r pkg`**.
The largest difference is that **`r pkg`** uses **ggplot2** as the graphing engine, rather than base R.
This means that **`r pkg`** relies on the data frame structure of **ggplot2**, which is rapidly becoming common for most applications.
Thus, **`r pkg`** is the "ggplot2 version" of **diagram**.


# DiagrammeR

The **DiagrammeR** package (https://rich-iannone.github.io/DiagrammeR/) is the most flexible diagramming package available for R.
However, it is primarily for acyclic diagrams (no feedbacks) and it requires specifying models in non-R syntax (DOT language).
It also requires knowledge of Graphviz.


