---
title: Comparison to other R packages
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comparison to other R packages}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

# avoid check where vignette filename must equal the vignette title
options(rmarkdown.html_vignette.check_title = FALSE)
```

```{r setup, include = FALSE}
pkg = 'flowdiagramr' #so we can easily switch names
library(pkg, character.only = TRUE)
```


## Overview

Here, we'll provide a brief discussion how **`r pkg`** differs from and can interact with other packages. 


## diagram

The [**diagram** package](https://cran.r-project.org/web/packages/diagram/index.html) is - as far as we are aware - most similar to **`r pkg`**.
The largest difference is that **`r pkg`** uses **ggplot2** as the graphing engine, rather than base R.
This means that **`r pkg`** relies on the data frame structure of **ggplot2**, which is rapidly becoming common for most applications.
Thus, **`r pkg`** is in some sense the "ggplot2 version" of **diagram**. 
In addition, the way **`r pkg`** is structured means that you can start creating your diagrams with the package commands, then switch to manual edits by adding ggplot2 code if needed.
This is not easily possible with **diagram**.


## DiagrammeR

The [**DiagrammeR** package](https://cran.r-project.org/web/packages/DiagrammeR/index.html) is the most flexible diagramming package available for R.
However, it is primarily for acyclic diagrams (no feedback or interaction terms) and it requires specifying models in non-R syntax (DOT language).
It also requires knowledge of Graphviz and is not quite ggplot2 compatible. We considered using **DiagrammeR** to get the diagrams we wanted, but the fact that a lot of models have feedback loops, which are not easily supported by **DiagrammeR** made us decide that we couldn't use it and had to create a new package.

## Others

There are other packages available that can draw graphs, e.g. [dagitty](https://cran.r-project.org/web/packages/dagitty/index.html) for directed acyclic graphs or [igraph](https://cran.r-project.org/web/packages/igraph/index.html) for network graphs. We are not aware of any package that can easly create the kind of flow diagrams that **`r pkg`** is intended to produce (otherwise we wouldn't have written the package).

If you know of other packages that are worth pointing out, please let us know.




