---
title: Modifying the ggplot2 code 
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Modifying the ggplot2 code}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7
)

# avoid check where vignette filename must equal the vignette title
options(rmarkdown.html_vignette.check_title = FALSE)
```

```{r setup, include = FALSE}
pkg = 'flowdiagramr' #so we can easily switch names
library(pkg, character.only = TRUE)
```

## Introduction

We assume you went through the previous vignettes that showed you how to use the functionality provided by **`r pkg`** to produce fairly good looking (in our biased opinion) diagrams. We believe (hope) that this often gives you a diagram you are happy with.

However, occasionally you might want to make additional tweaks beyond what **`r pkg`** can do. Fortunately for you, there is a way to do this, which is the topic of this vignette.



## Modifying the ggplot2 code

The most flexible approach to adjust diagrams is to modify the ggplot code that produces the diagram. To do this, you can retrieve the full ggplot2 code that creates the diagram and then edit it like you would any ggplot figure. We'll illustrate this approach with a few simple examples. 


### Example 1

We'll start with a very basic example. First, you define your model as usual. 

```{r}
# specify the model
variables = c("S","I","R")
flows = list(S_flows = c("n", "-b*S*I", "-m*S"), 
             I_flows = c("+b*S*I","-g*I", "-m*I"), 
             R_flows = c("g*I", "-m*R"))
sirmodel = list(variables = variables, flows = flows)
```

Next, you prepare the model

```{r}
diag_list <- prepare_diagram(sirmodel)
```


Once you have the diagram list ready, you can send it to `write_diagram`, a function that writes the full code needed to generate the diagram. 
You can specify the location and name for the file. If you don't provide those, it will be stored in the current working directory under the name `diagram_code.R`. 

```{r, eval= FALSE}
write_diagram(model_list = sirmodel1, filename = 'sirmodel1_diagram_code.R', always_overwrite = TRUE)
```
If things worked well, you should get a message telling you the location where the code was saved.
The code that is generated is completely self-contained, that means you can give it to anyone, who can run it and reproduce your model figure.

If you run the code that is generated, it will create the diagram structure as an object called `diagram_plot`. 
You can then plot or save that object (there is code at the bottom of the generated file for that).

For this example, here is what you get. 
```{r, eval = TRUE}
source('sirmodel1_diagram_code.R')
plot(diagram_plot)
```

Having all the code that generates the figure also allows you to do some fine-tuning. We provided hopefully enough comments in the code to make it clear what happens where. That said, to be able to tweak the diagram this way requires being familiar with creating figures in ggplot. 

Let's assume for this example that you want to add some text below the `I` compartment. 
In this case, you don't need to modify any of the existing code, just add a some code at the end. 
The following lines of code appended to the bottom of `sirmodel1_diagram_code.R`, just before the final `plot` statement, will do the trick:

```{r, eval=FALSE}
#######################
### LAST PART OF THE GENERATED CODE BEFORE ADDITION
#######################
if(with_grid == FALSE) {
  diagram_plot <- diagram_plot +
    theme_void()
} else {
  # The else here may seem silly, but otherwise the returned plot is NULL
  diagram_plot <- diagram_plot
}

#######################
### YOUR NEW ADDITION
#######################
# make a new data frame of text
text_df <- data.frame(  x = 3,  y = -1,
  lab = "Infected and Infectious"
)
diagram_plot <- diagram_plot +
  geom_text(data = text_df, aes(x = x, y = y, label = lab), size = 8)

#######################
### PART OF THE GENERATED CODE AFTER ADDITION
#######################
# These lines plot or save the generated diagram. 
# Uncomment them if you want to perform either action. 
# plot(diagram_plot) 
# ggsave('diagram_plot.png',diagram_plot)
```

Then save the new code, let's call it `sirmodel1_diagram_code_modified.R`.
Running the new code will add the wanted text to the diagram plot


```{r, eval = TRUE}
source('sirmodel1_diagram_code_modified.R')
plot(diagram_plot)
```



### Example 2

It is possible, and advisable, to first do modifications using the previously explained approaches. Those can be combined with this approach for editing the code. To that end, you can supply the `write_diagram` function all settings that you can send to `make_diagram`, and you can also supply your own modification of the `diagram_list` object. In the latter case, providing the model as input to `write_diagram` is not needed. You can still do so, the diagram will be stored inside the R code, but it will not be processed, instead your supplied `diagram_list` will be used.

Here is a full example showing how to use the different customization options for the SIR model.

Specify the model, including locations and variable names.

```{r}
varlabels = c("S","I","R")
flows = list(S_flows = c("-b*S*I"), 
             I_flows = c("+b*S*I","-g*I"), 
             R_flows = c("+g*I"))
varnames = c("Susceptible","Infected","Recovered")
varlocations = matrix(data = c("S", "", "R",
                               "", "I", "" ),nrow = 2, ncol = 3, byrow = TRUE)
sirmodel2 = list(varlabels = varlabels, flows = flows)
sirsettings2 = list(varlocations = varlocations)
```

Make and modify the `diagram_list` object by flipping the b*S*I interaction arrow to go below the S $\rightarrow$ I arrow. (There's no good reason to do that, other than to show an example.)

```{r}
diagram_list = prepare_diagram(sirmodel2, sirsettings2)
diagram_list$flows$ymin[3] = diagram_list$variables$ymin[2]
# diagram_list$flows$yend[3] = -1
diagram_list$flows$xlabel[3] = 1.5
diagram_list$flows$ylabel[3] = 0.5
diagram_list$flows$curvature[3] = -0.7
```

Create the diagram the usual way using `make_diagram`, while also supplying additional settings.


```{r}
#specify the settings separately so we can re-use them below
diagram_settings = list(var_fill_color = c("#6aa4c8", "#eb5600", "#1a9988"),
                        var_label_size = 14,
                        main_flow_color = "blue",
                        main_flow_linetype = "dotted")
sir2_diagram <- make_diagram(diagram_list, diagram_settings)
plot(sir2_diagram)
```

Alternatively, you can generate the full code that produces this plot as follows


```{r, eval= FALSE}
write_diagram(model_list = sirmodel2, 
              model_settings = sirsettings2,
              diagram_list = diagram_list,
              diagram_settings = diagram_settings,
              filename = 'sirmodel2_diagram_code.R',
              always_overwrite = TRUE)
```

If you run the code that is generated, you should see a figure like this

```{r, eval = TRUE}
source('sirmodel2_diagram_code.R')
plot(diagram_plot)
```

All of this you can get by using the previous methods of modification we explained. 

Let's assume that you are mostly happy with this figure, but you want to make another tweak.
This time, you want to (for some reason) turn the shape of the `I` box into a circle.

Here is how you modify the `sirmodel2_diagram_code.R` code to achieve this.


```{r, eval = FALSE}
#######################
### LAST PART OF THE GENERATED CODE BEFORE ADDITION
#######################
for(i in 1:nrow(variables)) {

#######################
### YOUR NEW ADDITION
#######################
if(i == 2) {
    diagram_plot <- diagram_plot +
      geom_point(
        data = variables[i, ],
        aes(x = labelx, y = labely),
        color = variables[i, "fill"],
        fill = variables[i, "fill"],
        shape = 21,
        size = 23
      )
  } else {
    diagram_plot <- diagram_plot +
      geom_rect(
        data = variables[i, ],
        aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
        color = variables[i, "color"],
        fill = variables[i, "fill"]
      )
  }
```


Now save the modified script (here We call it `sirmodel2_diagram_code_modified.R`), source it and plot the generated `diagram_plot` object to get this figure:

```{r, eval = TRUE}
source('sirmodel2_diagram_code_modified.R')
plot(diagram_plot)
```


### Example 3

Let's look at one more example. We'll use the predator-prey model again.

```{r}
# specify the model
varlabels = c("B","I")
flows = list(B_flows = c("+g*B*(1-B/bmax)", "-dB*B", "-k*B*I"),
             I_flows = c("+r*B*I", "-dI*I"))
ppmodel = list(varlabels = varlabels, flows = flows)
```

We won't do any changes to the `diagram_list` structure here, but we'll send a few custom `make_diagram` settings to the `write_diagram` code.

Here is first the plot using the `prepare_diagram` and make_diagram` functions. It produces as customized (somewhat ugly) diagram.

```{r}
diagram_list = prepare_diagram(ppmodel)
diagram_settings = list(
                          var_outline_color = "red",
                          var_fill_color = c("#6aa4c8", "green"),
                          var_label_color = "red",
                          var_label_size = 10,
                          main_flow_label_size = 4,
                          interaction_flow_size = 2,
                          main_flow_color = "green",
                          main_flow_linetype = 3)
pp_diagram <- make_diagram(diagram_list, diagram_settings)
plot(pp_diagram)
```

Now let's generate the code for this figure.

```{r, eval = FALSE}
# make diagram with adjusted settings
write_diagram(model_list = ppmodel, 
              diagram_settings = diagram_settings,
              filename = 'ppmodel_diagram_code.R',
              always_overwrite = TRUE)
```

Running the resulting code produces this diagram.

```{r, eval = TRUE}
source('ppmodel_diagram_code.R')
plot(diagram_plot)
```


Let's say that for some reason, you want the rectangles with rounded corners. I didn't know how to do that, but a quick online search [provided this solution](https://stackoverflow.com/questions/52820756/roundrectgrob-equivalent-for-geom-rect-in-ggplot). Note that this is somewhat advanced and hacky, but I figured I show it as an example to illustrate that at this point, you can do anything you like with the code to make whatever you want.

It's actually not much code editing. First, you need to install the [`statebins` R package](https://github.com/hrbrmstr/statebins). The version on CRAN works fine, so you can just use `install.packages('statebins')`. Then add `library('statebins')` at the top of your code file, underneath the other `library()` commands. 
Finally, find the `geom_rect` bit in your code (close to the top) and replace it with `statebins:::geom_rrect` (yes, you need the ::: in there since `geom_rrect` is not an exported function. If you have no idea what that means, that's ok.) Everything else you can keep unchanged. Save your code (I recommend using a different name so you don't accidentally overwrite it if you run your `write_diagram` code again). Now, if you run the code, it should produce this diagram.

```{r, eval = TRUE}
source('ppmodel_diagram_code_modified.R')
plot(diagram_plot)
```


## General suggestions

I think with those examples, you get the idea. At this point, the options are endless and you can change the code to get exactly what you want - at the expense of having to write/edit code yourself.

In general, we recommend implementing the easiest modifications first, even if you know you'll be editing the ggplot code at the end. Thus, first use the options described in the [basic modification vignette](./B_modify_diagrams_simple.html). Then make further adjustments by editing the `diagram_list` structure as described in the [input structure modification](./C_modify_input_structure.html) vignette. Once you have gotten as far as you can with these approaches, and still want to make further edits, write the ggplot code to a file and change the code.

Of course, once you have started manual editing of the code, things are not reversible, i.e. you have basically left the **`r pkg`** package infrastructure and are now adjusting the diagram on your own. You can of course always re-generate the code and copy your modifications over (as long as you remember to not overwrite your code.)



## Next steps

This concludes the main tutorials for the package. Several additional vignettes are available. One shows further model examples, another one discusses additional topics such as using **`r pkg`** together with our [modelbuilder package](https://ahgroup.github.io/modelbuilder/) and how to crop white space around the diagrams. Yet another vignette provides some examples on how to create flow diagrams for scenarios that are not based on compartmental models. It illustrates how basically anything for which you want to make a diagram consisting of nodes/boxes and flows/arrows can be created with **`r pkg`**. And one more vignette compares **`r pkg`** to similar R packages. You can access all those additional vignettes [from the package website](https://andreashandel.github.io/flowdiagramr/).









